Basic Parameter estimation of Binary Neutron Star Systems by the Advanced LIGO/Virgo Network

### 2013-06-11 doc/paper.tex
added bens comments

\begin{document}
\title{Basic Parameter estimation of Binary Neutron Star Systems by the Advanced LIGO/Virgo Network}
\author{Carl L. Rodriguez \altaffilmark{1}
%\email{rodriguez@u.northwestern.edu}
,Benjamin Farr \altaffilmark{1}
%\email{bfarr@u.northwestern.edu} 
Vivien Raymond \altaffilmark{1,2}
Will M. Farr \altaffilmark{1}
%\email{w-farr@northwestern.edu} 
Tyson Littenberg\altaffilmark{1}
%\email{tyson.littenberg@northwestern.edu}
Diego Fazi\altaffilmark{1}
%\email{d-fazi@northwestern.edu}
Vicky Kalogera\altaffilmark{1}}
%\email{vicky@northwestern.edu}

@@

Within the next five years, the Advanced LIGO/Virgo network will have reached a sensitivity sufficient to enable the routine
 detection of gravitational waves.  Beyond the initial detection, the scientific promise of these instruments relies the 
 parameter estimation techniques being currently developed.  The majority of this effort has been towards the detection
   and characterization of gravitational waves from compact binary coalescence, i.e. the coalescence of black holes and/or 
   neutron stars.  While several previous studies have focused on the parameter estimation abilities of advanced detectors, 
   the majority have relied on approximation techniques such as the Fisher Matrix.  Here, we report the statistical 
   uncertainties that will be achievable, in practice, using the parameter estimation techniques which will be applied to 
   detections of binary neutron star systems.  We find the recovery of the individual masses to be within fractionally within 
   10\% at the 65\% confidence level for equal-mass systems, and within 2\% for unequal-mass systems.
     We also report the average uncertainties on the sky-locations, luminosity distance, and orbital inclinations of randomly 
     distributed sources that can be achieved by different network configurations.
\end{abstract}

@@

 By the end of the decade, the field of gravitational-wave astrophysics will
 mature into an ideal tool for 
 exploring the gravitational side of the universe. Within the next few
 years, the first generation of gravitational-wave detectors capable 
 of regularly resolving astrophysical sources will come online \carl{[CITE]}.
 The Advanced LIGO and Virgo detectors will provide the first insights
 into the final moment of binary compact object mergers, including the 
 merger of binary neutron star systems.  Intense preparations are underway
 to characterize and extract as much physical 
 information as possible from these signals.

The mergers of binary neutron stars are expected to be the most common
source in the advanced detector era \carl{[CITE]}.   Models from stellar 
evolution suggest that the number of binary neutron star mergers within 
Advanced LIGO/Virgo's detection horizon could reach into the hundreds per 
year \carl{[CITE]}.  Although the peak sensitivity of ground-based detectors
is not focused on the frequency at which BNS systems merge, it could still
be possible to extract information about both strong field gravitational 
physics \carl{[CITE]} and the hydrodynamics of dense matter (e.g. the equation 
of state of nuclear matter) \carl{[CITE]}.  Furthermore, the observations of multiple BNS
systems will provide key insight into the evolution of binary systems in the field \carl{[CITE]},
and in globular clusters \carl{[CITE]}. As such, BNS systems will likely 
form the ``bread and butter'' of the compact binary coalescence detection effort in the coming 
years.

Of course one must distinguish between the detection of 
such events, and the precision measurement of the relevant physical parameters.
The detection of BNS systems will be performed with a matched filtering approach.  
By comparing the data stream with a bank of theoretical templates, the 
 time-series data can be searched for candidate signals at a sufficiently
  rapid rate to analyze months of data.  When the overlap between one of the template waveforms and the 
detector output is sufficiently large (97\%), a detection candidate is found.  However,
the parameter space of these signals can be highly degenerate, with several 
locations in parameter space corresponding to nearly identical waveforms.  In 
order to fully realize the science potential of these instruments, we must 
perform a full exploration of the parameter space for each detection.  By 
analyzing the parameter space of each candidate with a Bayesian inference technique 
 we will be able to make precise, scientifically meaningful
statements about the physics of these detectible systems.  To 
that end, we employ a Markov-Chain Monte Carlo sampling algorithm, 
\texttt{lalinference\_mcmc}, included in the LIGO Application Library parameter
estimation code, \texttt{LALInference}, to analyze the parameter space of binary neutron star systems.

By employing the full parameter-estimation machinery that will eventually be
used in the Advanced LIGO/Virgo era, our results give the first estimate at the 
realistic capabilities of advanced ground-based detectors to characterize BNS systems.
  Until recently, the majority of studies have employed the Fisher matrix formalism which was 
first adapted for gravitational-wave parameter estimation by \citep{FinnDetection}.
While each of these studies (\cite{PoissonWill,CutlerFlanagan,ArunPE}, among 
many others) have pointed out the limitations 
and flaws of the Fisher Information Matrix, there has been no study which investigates
the BNS parameter estimation capabilities of Advanced LIGO/Virgo via the techniques
that will eventually be employed.  The recent work of \cite{Vallisneri} and our 
own work \citep{Inadequacies} has demonstrated that the Fisher matrix cannot even
be treated as a lower bound on the standard deviations of certain parameters
(particularly the masses) measurable in gravitational-wave detectors.

In this paper, we perform a systematic study of the statistical uncertainties with
which the Advanced LIGO/Virgo network will be able to measure the basic parameters of 
binary neutron stars.  In Section \ref{PEsection}, we describe
the machinery of our parameter estimation code, \texttt{LALInference}, and the associated
MCMC sampler, \texttt{lalinference\_mcmc}, as well as the frequency-domain 
gravitational-wave template we employ.   In Section \ref{resultsSection}, we qualitatively 
analyze the posterior probability density functions for BNS systems with different masses 
and extrinsic parameters.  The results are divided into three parameter sets of interest: the
recovery of the mass parameters (Section \ref{massSection}), the recovery of the orbital
inclination and luminosity distance (Section \ref{idSection}), and the localization of sources 
in the sky (Section \ref{skySection}).   Finally we provide quantitative 1-dimensional confidence
 intervals on the parameter recovery in Section \ref{ciSection}.  These results, contained
 in Tables \ref{ciTableIntrinsic} and \ref{ciTableExtrinsic}, are intended as a reference for the 
 optimal BNS parameter estimation capabilities of Advanced LIGO/Virgo.
   We assume that $G=c=1$ throughout.

@@

We begin by introducing a Bayesian formalism for parameter estimation. 
%Before offering a brief derivation of the FIM, a few preliminary
%techniques must be introduced.  First,
We assume that the time-domain signal in a gravitational-wave
network can be written as a combination of a gravitational
waveform $h_0$ and the noise of the detector $n$.  We further assume
that this noise is stationary and Gaussian with zero mean.  Therefore,
the detector output is simply
\begin{equation}
s = n + h_0 .
\label{SignalAddition}
\end{equation}
Since our noise model is Gaussian, we can write the
probability of a specific signal realization $s$ as proportional to the probability
that our residual is Gaussian distributed once the waveform has been subtracted
\begin{align}
  p(s | \thpara) &\propto \exp\left[-\frac{1}{2}\left<n|n
    \right>\right] \nonumber \\ &= \exp\left[-\frac{1}{2}\left < s -
    h(\thpara) | s-h(\thpara)\right >\right] ,
  \label{likelihood}
\end{align}
where $\thpara$ is the set of parameters for our template
waveforms, and $p(s | \thpara)$ is the likelihood of the signal $s$ given the parameters $\thpara$.
  The inner product, $\left< ~|~ \right> $, is defined using the noise
spectrum of the detectors as
\begin{equation}
  \left<a|b\right> \equiv 4 \Re \int \frac{\tilde{a}(f)\tilde{b}^*(f)}{S_n(f)} df ,
  \label{innerProduct}
\end{equation}
where $S_n(f)$ is the one-sided power spectral density as a function
of frequency, and $\tilde{a}(f)$ and $\tilde{b}(f)$ are the Fourier
transforms of the time-domain signals $a(t)$ and $b(t)$.

@@

We are interested in the posterior $p(\thpara | s)$ as it encodes information about
prior state of knowledge about the problem.  For instance, since it is assumed that compact binaries
form homogeneously in the local universe, we can safely assume that the population of
sources will be distributed homogeneously in volume, or evenly in $D^2$. 
 Our prior information can come from mathematical limits in the parameter space or
from our \textit{a priori} knowledge of astrophysical systems.  We employ priors which are

@@

If we pick a set of parameters $\thpara$ such that
$h(\thpara) = h_0$, then the posterior \eqref{posterior}
will be near a global maximum; however, the presence of noise will in
general deflect the maximum of our posterior away from the value at
$\thpara_0$. That is, in the presence of noise, there is
no guarantee that the maximum-likelihood of our signal corresponds to
the true parameters of the system.

@@

The LIGO Algorithm Library parameter estimation code, \texttt{LALInference}, is designed as 
a unified framework for gravitational-wave parameter estimation.  By using a common setup for 
waveform generation, PSD estimation, data handling, and other associated techniques from gravitational-wave parameter estimation, \texttt{LALInference} allows the implementation of multiple parameter samplers in using identical parameter setups, including Nested Sampling (\texttt{lalinference\_nest})\carl{[CITE]} and Markov-Chain Monte Carlo (\texttt{lalinference\_mcmc}).  We elect to use the MCMC sampler for this study.  \texttt{lalinference\_mcmc} is based upon the previously described code, 
\texttt{SpinSpiral} \citep{spinspiral2009, spinspiral2010}.  The MCMC employs a Metropolis-Hastings sampling algorithm, which can be described as follows \citep{Gilks99}:

@@

The above procedure is designed to record a chain of samples whose
distribution is $p\left(\thpara|s\right)$.  By drawing a sufficient ($\sim 1000$) number
 of effective samples from the posterior, the chains traces out the functional form of the posterior,
 gathering more samples from regions with high posterior probability.  Depending on
the proposal distribution, $q$, the convergence (mixing) of the chain
may be rapid or slow.  We employ multiple optimization techniques,
including both specially-crafted $q$ and parallel tempering, to ensure
adequate mixing of the Markov Chains throughout our parameter space.
The details of the algorithm can be found in
\citep{Sluys08,spinspiral2009, spinspiral2010}.  Both samplers were tuned and 
developed during the last science run of the Initial LIGO/Virgo network.  
A description of the parameter estimation capabilities of these two samplers with respect to
real gravitational-wave data, as well as a more detailed description of the algorithms and
checks for convergence, can be found in \cite{S6PE}.

@@

We use a frequency domain waveform accurate up to $3.5$
post-Newtonian (pN) order in phase and amplitude.  We restrict ourselves to
quasi-circular, non-spinning waveforms as a simplifying assumption.  The standard
form of our waveform model, known as the \textit{TaylorF2}
approximant, is calculated via the stationary-phase approximation.
  In this setup, the gravitational-wave
amplitude is given by
\begin{equation}
\tilde{h}(f) = a(t_f) e^{i \psi(f)},
\label{amplitude}
\end{equation}
where $a(t_f)$ is the amplitude evaluated at a stationary-phase reference point, which to lowest order
takes the form $A \propto f^{-7/6} \chmass^{5/6}\Theta(\text{angle})/D$, $D$ is
the luminosity distance of the binary, and $\psi(f)$ is the pN phase.
$\Theta(\text{angle})$ is a function of the orbital orientation with
respect to the detector network in terms of the sky position, orbital
inclination, and the wave polarization.  In addition to the total
mass, $M_{tot}\equiv M_1+M_2$, it is convenient to work with the
\textit{mass ratio} and \textit{chirp mass}, defined by
\begin{equation}
  q\equiv M_1/M_2~~~~\text{and}~~~~\chmass = (M_1 M_2)^{3/5} M^{-1/5},
  \label{eqRatioCM}
\end{equation}
respectively.\footnote{Most gravitational-wave literature instead uses the \textit{symmetric mass ratio}, defined as $\eta \equiv M_1 M_2 / M^2$.  We elect to use $q$ as it is more physically intuitive, and the prior $p(q)$ is uniform when the component
mass priors are uniform.}  The stationary phase then becomes an expansion in the 
Newtonian orbital velocity, $v=(\pi M f)^{1/3}$, 
\begin{equation}
\psi(f) = 2 \pi f t_c - \phi_0 + \frac{\pi}{4} + \frac{3}{128}\left(\frac{M}{\chmass}\right)^{5/3}\sum^{n}_{k=0}\alpha_{k}v^{k-5}
\label{phase}
\end{equation}

@@

To perform the integral defined in \eqref{innerProduct}, we used as our power-spectral
density the best estimate for a high-power, zero-detuning configuration of Advanced 
LIGO, provided by the LIGO Scientific Collaboration.  Both the noise curve and technical 
reports describing it can be found in \cite{ADVLIGONoise}.
We consider two configurations of the advanced detector
network: a three-detector configuation consistanting of the two LIGO sites 
(in Hanford, WA and Livingston, LA) and the Virgo site (in Pisa, Italy), and a 
four-detector configuration with adds the proposed LIGO-India detector
(in Chitradurga, KA).  For simplicity, we assume each detector to be 
operating at the Advanced LIGO sensitivity.

@@

\begin{equation}
  \pi f_{\text{ISCO}} = \frac{1}{6^{3/2}M}.
  \label{ISCOFrequency}
\end{equation}

We consier the mass parameters ($\chmass,q$) and the phasing parameters,
($\phi_0,t_c$), to be the intrinsic parameters of a gravitational-wave signal.  There 
are an additional 5 extrinsic parameters, independent of the pN orbital phase, which can
modulate the amplitude of the signal in an independent fashion for each detector.  
Combining these leads to our 9-dimensional parameter space for non-spinning systems:
\begin{equation}
\thpara =  (\chmass, q, \phi_0,t_c,D,\iota,\psi,\alpha,\delta)
\label{parameterspace}
\end{equation}
where to summarize

\begin{itemize}
\item $\chmass$ is the chirp mass,
\item $q$ is the mass ratio,
\item $\phi_0$ and $t_c$ are the chirp phase and chirp time, arbitrary phasing parameters,
\item $D$ is the luminosity distance to the binary,
\item $\iota$ is the orbital inclination, 
\item $\psi$ the gravitational-wave polarization, and
\item $\alpha$ and $\delta$ the right ascension and declination of the
source on the sky.
\end{itemize}
  Since the wave amplitude depends on the orientation of the
binary with respect to each detector, most of the information about these extrinsic
parameters comes from two sources: the time-of-arrival triangulation of 
the source locations, and the relative amplitudes within the detector network.

We define the signal-to-noise ratio (SNR) of a gravitational wave in a
single detector as 
\begin{equation}
  \rho \equiv \frac{4}{\sigma} \int^{\infty}_{0}\frac{| \tilde{s}(f)\tilde{h}(f)|}{S_{n}(f)}df
  \label{formalSNR}
\end{equation}
where $\rho$ is the SNR and $\tilde s(f)$ and $\tilde{h}^{*}(f)$ are
the frequency-domain signal and template, respectively, and the normalization $\sigma$ is
given by
\begin{equation}
  \sigma^2 = 4\int^{\infty}_{0}\frac{| \tilde{h}(f)|^2}{S_n(f)}df.
  \label{SNRnorm}
\end{equation}

When dealing with network of gravitational-wave detectors, the SNRs of the individual detectors add in quadrature.  That is, the network SNR for a detection is

@@

For this study, we consider four seperate populations of BNS systems, with component mass 
combinations of $1M_{\odot}/1M_{\odot}$, $1.4M_{\odot}/1.4M_{\odot}$, $1M_{\odot}/2.5M_{\odot}$, 
and $2.5M_{\odot}/2.5M_{\odot}$.  Each population consisted of 40 signals distributed randomly
in sky location, polarization, inclination, time-of-arrival, and coalescence phase.  The luminosity 
distance, $D$, was selected to provide a network SNR of $\rho_{\text{network}}=20$ for each source.
While unphysical from an astrophysical perspective (binary sources should be 
distributed evenly in comoving volume), this choice allows us to explore the parameter estimation
in the context of optimal detection candidates.

@@

We find that there is virtually no difference between the mass PDFs of our different injected signals
within each mass bin.  While this may initially seem disconcerting, it is to be expected.  Recall that of the BNS signals were injected with a network SNR of 20 into a zero-noise detector realization.  Furthermore, note that the mass parameters are the only two which directly effect the phase of the TaylorF2 waveform \eqref{phase}.  Therefore, as long as the injected mass parameters are identical within each mass system, and as long as the injected SNR is identical, the amount of recoverable information in each mass parameter should be almost identical, and for a sufficiently converged MCMC chain, the recovered posterior PDF should be identical.

@@

To condense the results into a single figure of merit, we average the 1-dimensional mass PDFs into a single 
posterior probability for system.  See Figure~\ref{metaMassPDFs}.  Notice how when averaged and
 normalized to the injected values, the recovery of the component masses depends only on the mass ratio.  
 Furthermore, for systems of equal masses, the posterior barely extends beyond 15\% of the injected values.  
 If one assumes that the threshold between black
   holes and neutron stars lies at approximately $3M_{\odot}$, then it might be assumed that
   the mass recovery would allow one to discriminate between black holes and neutron stars.

Unfortunately, this will not be feasible in practice.  We have neglected the spin of the binaries, which will be highly
correlated to the masses.  This coupling means that a higher-massed spinning black hole systems can produce
 waveforms very similar to non-spinning, low-mass, making it impossible to discern between non-spinning neutron stars
 and spinning, more massive black holes.  However, the situation is not hopeless: if the spins are misaligned, the spin 
 vectors will couple to the orientation of the binary (encoded in the three angles $\phi_0$, $\iota$, and $\psi$) via
 relativistic precession.  It remains to be seen if using fully spinning waveforms will make it possible for Advanced\
  LIGO/Virgo to discern binary neutron stars from their spinning black hole counterparts.

In Table \ref{ciTableIntrinsic}, we list the 65\% fractional confidence intervals for the mass parameters.  We find that the 
component masses for equal mass systems can be isolated to between 6\% and 9\% fractional uncertainty at the 65\% confidence interval.  This value drops to less than 2\$ for the components of the unequal mass systems.  Again, this 
neglects the effects of spin which can substantially increase these values.

@@

\section{Confidence Intervals}
\label{ciSection}

>>>>

\begin{document}
\title{Basic Parameter estimation of Binary Neutron Star Systems by the Advanced LIGO/Virgo Network}
\author{Carl L. Rodriguez \altaffilmark{1}
%\email{rodriguez@u.northwestern.edu}
Benjamin Farr \altaffilmark{1}
%\email{bfarr@u.northwestern.edu} 
Vivien Raymond \altaffilmark{1,2}
Will M. Farr \altaffilmark{1}
%\email{w-farr@northwestern.edu} 
Tyson Littenberg\altaffilmark{1}
%\email{tyson.littenberg@northwestern.edu}
Diego Fazi\altaffilmark{1}
%\email{d-fazi@northwestern.edu}
Vicky Kalogera\altaffilmark{1}}
%\email{vicky@northwestern.edu}

@@

Within the next five years, the Advanced LIGO/Virgo network will have reached a sensitivity sufficient to enable the routine
 detection of gravitational waves.  Beyond the initial detection, the scientific promise of these instruments relies on the ability to perform parameter estimation. The majority of this effort has been towards the detection
   and characterization of gravitational waves from compact binary coalescence, e.g. the coalescence of binary 
   neutron stars.  While several previous studies have focused on the parameter estimation abilities of advanced detectors, 
   the majority have relied on approximation techniques such as the Fisher Matrix.  We report the statistical 
   uncertainties that will be achievable, in practice, using the full parameter estimation machinery developed by the  
   LIGO/Virgo Collaboration.  We find the recovery of the individual masses to be fractionally within 
   10\% at the 65\% credible intervals for equal-mass systems, and within 2\% for unequal-mass systems.
     We also report the average uncertainties on the sky-locations, luminosity distance, and orbital inclinations of randomly 
     distributed sources that can be achieved by different network configurations.
\end{abstract}

@@

Within the next few
 years, the first generation of gravitational-wave detectors capable 
 of regularly resolving astrophysical sources will come online \citep{AdvLIGO,AdvVirgo}.
 The Advanced LIGO and Virgo detectors will provide the first insights
 into the final moment of binary compact object mergers, including the 
 merger of binary neutron star systems.  Intense preparations are underway
 to characterize and extract as much physical 
 information as possible from these signals.

The mergers of binary neutron stars (BNS) are expected to be the most common
compact binary source in the advanced detector era.   Models from stellar 
evolution and observations of binary pulsars suggest that the number of binary neutron star mergers within 
Advanced LIGO/Virgo's detection horizon could reach into the hundreds each 
year, with an expected mean of $\sim 40$ per year \citep{RatesPaper}. 
Although the peak sensitivity of ground-based detectors
is not focused on the frequency at which BNS systems merge, it could still
be possible to extract information about both strong field gravitational 
physics \carl{[CITE]} and the hydrodynamics of dense matter (e.g. the equation 
of state of nuclear matter) \carl{[CITE]}.  Furthermore, the observations of multiple BNS
systems will provide key insight into the evolution of binary systems in the field \citep{VickyRates,KimRates}
. As such, BNS systems will likely 
form the ``bread and butter'' of the compact binary coalescence detection effort in the coming 
years.

Of course one must distinguish between the detection of 
such events, and the precision measurement of the relevant physical parameters.
The detection of BNS systems will be performed with a matched filtering approach.  
By comparing the data stream with a bank of theoretical templates, the 
 time-series data can be searched for candidate signals at a sufficiently
  rapid rate to analyze months of data.  When the overlap between one of the template waveforms and the 
detector output is sufficiently large (97\%), a detection candidate is found.  However,
the parameter space of these signals can be highly degenerate, with several 
locations in parameter space corresponding to nearly identical waveforms.  In 
order to fully realize the science potential of these instruments, we must 
perform a full exploration of the parameter space for each detection.  By 
analyzing the parameter space of each candidate with a Bayesian inference technique, 
 we will be able to make precise, scientifically meaningful
statements about the physics of these systems.  To 
that end, we employ a Markov-Chain Monte Carlo sampling algorithm, 
\texttt{lalinference\_mcmc}, included in the LIGO Application Library parameter
estimation code, \texttt{LALInference}, to analyze the parameter space of BNS systems.

By employing the full parameter-estimation machinery that will eventually be
used in the Advanced LIGO/Virgo era, our results give the first estimates at the 
realistic capabilities of advanced ground-based detectors to characterize BNS systems.
  Until recently, the majority of studies have employed the Fisher matrix formalism which was 
first adapted for gravitational-wave parameter estimation by \citep{FinnDetection}.
While each of these studies \citep{PoissonWill,CutlerFlanagan,ArunPE} have pointed out the limitations 
and flaws of the Fisher Information Matrix, there has been no study which investigates
the BNS parameter estimation capabilities of Advanced LIGO/Virgo using the techniques
that will eventually be employed.  The recent work of \cite{Vallisneri} and our 
own work \citep{Inadequacies} has demonstrated that the Fisher matrix cannot even
be treated as a lower bound on the standard deviations of certain parameters
(particularly the masses) measurable in gravitational-wave detectors.

In this paper, we perform a systematic study of the statistical uncertainties with
which the Advanced LIGO/Virgo network will be able to measure the basic parameters of 
BNS mergers.  In Section \ref{PEsection}, we describe
the machinery of our parameter estimation code, \texttt{LALInference}, and the associated
MCMC sampler, \texttt{lalinference\_mcmc}, as well as the frequency-domain 
gravitational-wave template we employ.   In Section \ref{resultsSection}, we qualitatively 
analyze the posterior probability density functions for BNS systems with different masses 
and extrinsic parameters.  The results are divided into three parameter sets of interest: the
recovery of the mass parameters (Section \ref{massSection}), the recovery of the orbital
inclination and luminosity distance (Section \ref{idSection}), and the localization of sources 
on the sky (Section \ref{skySection}).   Finally we provide quantitative 1-dimensional confidence
 intervals on each parameter in Section \ref{ciSection}.  These results, contained
 in Tables \ref{ciTableIntrinsic} and \ref{ciTableExtrinsic}, are intended as a reference for the 
 optimal BNS parameter estimation capabilities of Advanced LIGO/Virgo.
   We assume that $G=c=1$ throughout.

@@

We begin by introducing a Bayesian formalism for parameter estimation. 
%Before offering a brief derivation of the FIM, a few preliminary
%techniques must be introduced.  First,
We assume that the time-domain signal in a gravitational-wave
network can be written as a combination of a gravitational
waveform $h_0$ and the noise of the detector $n$.  We further assume
that this noise is stationary and Gaussian with zero mean.  Therefore,
the detector output is simply
\begin{equation}
s = n + h_0 .
\label{SignalAddition}
\end{equation}
Since the noise model is Gaussian, we can write the
probability of a specific signal realization $s$ as proportional to the probability
that the residual is Gaussian distributed once the waveform has been subtracted
\begin{align}
  p(s | \thpara) &\propto \exp\left[-\frac{1}{2}\left<n|n
    \right>\right] \nonumber \\ &= \exp\left[-\frac{1}{2}\left < s -
    h(\thpara) | s-h(\thpara)\right >\right] ,
  \label{likelihood}
\end{align}
where $\thpara$ is the set of parameters for the template
waveforms, and $p(s | \thpara)$ is the likelihood of the signal $s$ given the parameters $\thpara$.
  The inner product, $\left< ~|~ \right> $, is defined using the noise
spectrum of the detectors as
\begin{equation}
  \left<a|b\right> \equiv 4 \Re \int \frac{\tilde{a}(f)\tilde{b}^*(f)}{S_n(f)} df ,
  \label{innerProduct}
\end{equation}
where $S_n(f)$ is the one-sided power spectral density (PSD) as a function
of frequency, and $\tilde{a}(f)$ and $\tilde{b}(f)$ are the Fourier
transforms of the time-domain data $a(t)$ and $b(t)$.  If we pick a set of parameters $\thpara$ such that
$h(\thpara) = h_0$, then the posterior \eqref{posterior}
will be near a global maximum; however, the presence of noise will in
general deflect the maximum of our posterior away from the value at
$\thpara_0$. That is, in the presence of noise, there is
no guarantee that the maximum-likelihood of our signal corresponds to
the true parameters of the system.

@@

We are interested in the posterior $p(\thpara | s)$ as it encodes information about the
prior state of knowledge about the problem in addition to the likelihood of the signal.
  For instance, since it is assumed that compact binaries
form homogeneously in the local universe observable by advanced detectors,
 we can safely assume that the population of
sources will be distributed homogeneously in volume, or evenly in $D^2$. 
 Our prior information can come from mathematical limits in the parameter space or
from our \textit{a priori} knowledge of astrophysical systems.  We employ priors which are

@@

\noindent  The mass priors listed here are the generic priors used in follow-up analysis, not specifically for BNS systems.  While we could have used a prior with a lower component mass cutoff for the present study, in practice the higher prior boundary does not effect our current analysis.

@@

The LIGO Algorithm Library Bayesian inference code, \texttt{LALInference}, is designed as 
a unified framework for gravitational-wave parameter estimation.  By using a common setup for 
waveform generation, PSD estimation, data handling, and other associated techniques from gravitational-wave parameter estimation, \texttt{LALInference} allows the implementation of multiple samplers of the parameter space, including Nested Sampling (\texttt{lalinference\_nest}, described in \cite{nestedsampling2010}) and Markov-Chain Monte Carlo (\texttt{lalinference\_mcmc}).  We elect to use the MCMC sampler for this study.  \texttt{lalinference\_mcmc} is based upon the previously described code, 
\texttt{SpinSpiral} \citep{spinspiral2009, spinspiral2010}.  The MCMC employs a Metropolis-Hastings sampling algorithm, which can be described as follows \citep{Gilks99}:

@@

The above procedure is designed to record a chain of samples whose
distribution is $p\left(\thpara|s\right)$.  By drawing a sufficient ($\sim1000$) number
 of effectively independent samples from the posterior, the chains traces out the functional form of the posterior,
 gathering more samples from regions with high posterior probability density.  Depending on
the proposal distribution, $q$, the convergence (mixing) of the chain
may be rapid or slow.  We employ multiple optimization techniques,
including both specially-crafted jump proposals ($q$) and parallel tempering, to ensure
adequate mixing of the Markov Chains throughout our parameter space.
Both samplers were tuned and 
developed during the last science run of the Initial LIGO/Virgo network.  
A description of the parameter estimation capabilities of these two samplers with respect to
real interferometer data, as well as a more detailed description of the algorithms and
checks for convergence, can be found in \cite{S6PE}.

@@

We use a frequency-domain waveform accurate up to $3.5$
post-Newtonian (pN) order in phase and 3 pN order in amplitude.  We restrict ourselves to
quasi-circular, non-spinning waveforms as a simplifying assumption.  The standard
form of our waveform model, known as the \textit{TaylorF2}
approximant, is calculated via the stationary-phase approximation.
  In this setup, the gravitational-wave
amplitude is given by
\begin{equation}
\tilde{h}(f) = a(t_f) e^{i \psi(f)},
\label{amplitude}
\end{equation}
where $a(t_f)$ is the amplitude evaluated at a stationary-phase reference point, which to lowest order
takes the form $a(t_f) \propto f^{-7/6} \chmass^{5/6}\Theta(\text{angle})/D$, where $D$ is
the luminosity distance of the binary, and $\psi(f)$ is the pN phase.
$\Theta(\text{angle})$ is a function of the orbital orientation with
respect to the detector network in terms of the sky position, orbital
inclination, and the wave polarization.  In addition to the total
mass, $M_{tot}\equiv M_1+M_2$, it is convenient to work with the
mass ratio and chirp mass, defined by
\begin{equation}
  q\equiv M_1/M_2~~~~\text{and}~~~~\chmass = (M_1 M_2)^{3/5} M_{tot}^{-1/5},
  \label{eqRatioCM}
\end{equation}
respectively.\footnote{Most gravitational-wave literature instead uses the \textit{symmetric mass ratio}, defined as $\eta \equiv M_1 M_2 / M^2$.  We elect to use $q$ as it is more physically intuitive, and the prior $p(q)$ is uniform when the component
mass priors are uniform.}  By convention, $M_1 \geq M_2$.  The stationary phase then becomes an expansion in the 
Newtonian orbital velocity, $v=(\pi M_{tot} f)^{1/3}$, 
\begin{equation}
\psi(f) = 2 \pi f t_c - \phi_0 + \frac{\pi}{4} + \frac{3}{128}\left(\frac{M_{tot}}{\chmass}\right)^{5/3}\sum^{n}_{k=0}\alpha_{k}v^{k-5}
\label{phase}
\end{equation}

@@

To perform the integral defined in \eqref{innerProduct}, we used as our power-spectral
density the best estimate for a high-power, zero-detuning configuration of Advanced 
LIGO, provided by the LIGO Scientific Collaboration.  Both the noise curve and technical 
reports describing it can be found in \cite{ADVLIGONoise}.
We consider two configurations of the advanced detector
network: a three-detector configuation consistanting of the two LIGO sites 
(in Hanford, WA and Livingston, LA) and the Virgo site (in Pisa, Italy), and a 
four-detector configuration that adds the proposed LIGO-India detector
(in Chitradurga, KA).  For simplicity, we assume each detector to be 
operating at the Advanced LIGO sensitivity.

@@

\begin{equation}
  \pi f_{\text{ISCO}} = \frac{1}{6^{3/2}M_{tot}}.
  \label{ISCOFrequency}
\end{equation}

We consier the mass parameters ($\chmass,q$) and the phasing parameters,
($\phi_0,t_c$), to be the intrinsic parameters of a gravitational-wave signal.  There 
are an additional 5 extrinsic parameters, independent of the pN orbital phase, which can
modulate the amplitude of the signal in an independent fashion for each detector.  
Combining these leads to our 9-dimensional parameter space for non-spinning systems,
\begin{equation}
\thpara =  (\chmass, q, \phi_0,t_c,D,\iota,\psi,\alpha,\delta),
\label{parameterspace}
\end{equation}
where

\begin{itemize}
\item $\chmass$ is the chirp mass,
\item $q$ is the mass ratio,
\item $\phi_0$ and $t_c$ are the chirp phase and chirp time, arbitrary phasing parameters,
\item $D$ is the luminosity distance to the binary,
\item $\iota$ is the orbital inclination (the angle between the orbital angular momentum and the line of sight), 
\item $\psi$ is the gravitational-wave polarization, and
\item $\alpha$ and $\delta$ are the right ascension and declination of the
source on the sky.
\end{itemize}
  Since the wave amplitude depends on the orientation of the
binary with respect to each detector, most of the information about these extrinsic
parameters comes from two quantities: the time-of-arrival triangulation of 
the source locations, and the relative amplitudes within the detector network.

We define the signal-to-noise ratio (SNR) of a gravitational wave in a
single detector as 
\begin{equation}
  \rho \equiv \frac{4}{\sigma} \int^{\infty}_{0}\frac{| \tilde{s}(f)\tilde{h}(f)|}{S_{n}(f)}df,
  \label{formalSNR}
\end{equation}
where $\rho$ is the SNR and $\tilde s(f)$ and $\tilde{h}^{*}(f)$ are
the frequency-domain data and template, respectively, and the normalization $\sigma$ is
given by
\begin{equation}
  \sigma^2 = 4\int^{\infty}_{0}\frac{| \tilde{h}(f)|^2}{S_n(f)}df.
  \label{SNRnorm}
\end{equation}

When dealing with network a of gravitational-wave detectors, the SNRs of the individual detectors add in quadrature.  That is, the network SNR for a detection is

@@

For this study, we consider four seperate populations of BNS systems, with component mass 
combinations of $1M_{\odot}/1M_{\odot}$, $1.4M_{\odot}/1.4M_{\odot}$, $1M_{\odot}/2.5M_{\odot}$, 
and $2.5M_{\odot}/2.5M_{\odot}$.  Each population consisted of 40 signals distributed randomly
in sky location, polarization, inclination, time-of-arrival, and coalescence phase.  The luminosity 
distance, $D$, was selected to provide a network SNR of $\rho_{\text{network}}=20$ for each source.
While unphysical from an astrophysics perspective (binary sources should be 
distributed evenly in comoving volume), this choice allows us to explore the parameter estimation
in the context of optimal detection candidates.

@@

We find that there is virtually no difference between the mass PDFs of our different injected signals
within each mass bin.  While this may initially seem disconcerting, it is to be expected.  Recall that all of the BNS signals were injected with a network SNR of 20 into a zero-noise detector realization.  Furthermore, note that the mass parameters are the only two which directly effect the phase of the TaylorF2 waveform \eqref{phase}.  Therefore, as long as the injected mass parameters are identical within each mass system, and as long as the injected SNR is identical, the amount of recoverable information in each mass parameter should be almost identical, and for a sufficiently converged MCMC chain, the recovered posterior PDF should be identical.

@@

To condense the results into a single figure of merit, we average the 1-dimensional mass PDFs into a single 
posterior probability for the system.  See Figure~\ref{metaMassPDFs}.  Notice how when averaged and
 normalized to the injected values, the recovery of the component masses depends only on the mass ratio.  
 Furthermore, for systems with equal component masses, the posterior barely extends beyond 15\% of the injected values.  
 If one assumes that the threshold between black
   holes and neutron stars lies at approximately $3M_{\odot}$, then it might be assumed that
   the mass recovery would allow one to discriminate between black holes and neutron stars.

Unfortunately, this will not be feasible in practice.  We have neglected the spin of the binaries, which will be highly
correlated to the masses.  This coupling means that a higher-massed spinning black hole systems can produce
 waveforms very similar to non-spinning, low-mass systems, making it extremely difficult to discern between non-spinning 
 neutron stars
 and spinning, more massive black holes \citep{Baird2013,Hannam2013}.  However, the situation is not hopeless: if the spins are misaligned, the spin 
 vectors will couple to the orientation of the binary (encoded in the three angles $\phi_0$, $\iota$, and $\psi$) via
 relativistic precession.  It remains to be seen if using fully spinning waveforms will make it possible for Advanced\
  LIGO/Virgo to discern binary neutron stars from their spinning black hole counterparts.

In Table \ref{ciTableIntrinsic}, we list the 65\% fractional credible intervals for the mass parameters.  We find that the 
component masses for equal mass systems can be isolated to between 6\% and 9\% fractional uncertainty at the 65\% credible interval.  This value drops to less than 2\% for the components of the unequal mass systems.  Again, this 
neglects the effects of spin which can substantially increase these values.

@@

\section{Credible Intervals}
\label{ciSection}
